---
title: "Computing and Extracting SPEI"
output: html_notebook
---

This notebook computes the Standardized Precipitation-Evapotranspiration Index (SPEI) at 1-, 3-, 6-, and 12-month time scales from CRU TS 4.03 global climate data, then extracts the South Africa subset and saves each time scale as a CSV for downstream analysis. (Incorporates former script 2.)

```{r libraries}
library(SPEI)
library(ncdf4)
library(snowfall)
library(Hmisc)
```

## SPEI Computation

The `spei.nc` function reads global precipitation and potential evapotranspiration netCDF files, computes SPEI using parallelised grid-cell processing, and writes the result to a new netCDF file. The computation loop runs at all four time scales; this step is slow (potentially several hours) and only needs to be run once.

```{r compute-spei}
spei.nc <- function(sca, inPre, outFile, inEtp=NA, title=NA, comment=NA,
                    inMask=NA, block=18, tlapse=NA) {

  # Read data
  pre.nc <- nc_open(inPre)
  if (!is.na(inEtp)) {
    etp.nc <- nc_open(inEtp)
    isSPEI <- TRUE
  } else {
    isSPEI <- FALSE
  }

  if (!is.na(inMask)) {
    mask.nc <- nc_open(inMask)
    isMask <- TRUE
  } else {
    isMask <- FALSE
  }

  # Create variable and output file
  if (!is.na(tlapse[1])) {
    spi.dim.tme <- ncdim_def(pre.nc$dim$time$name, pre.nc$dim$time$units,
                             pre.nc$dim$time$vals[tlapse[1]:tlapse[2]])
  } else {
    spi.dim.tme <- ncdim_def(pre.nc$dim$time$name, pre.nc$dim$time$units,
                             pre.nc$dim$time$vals)
  }
  if (isSPEI) {
    nam <- 'spei'
    longnam <- 'Standardized Precipitation-Evapotranspiration Index'
  } else {
    nam <- 'spi'
    longnam <- 'Standardized Precipitation Index'
  }
  out.nc.var <- ncvar_def(
    name=nam,
    units='z-values',
    dim=list(pre.nc$dim$lon, pre.nc$dim$lat, spi.dim.tme),
    missval=1.e30,
    longname=longnam,
    prec='float',
    compression=9)
  out.nc <- nc_create(outFile, out.nc.var)
  ncatt_put(out.nc,0,'Title',title)
  ncatt_put(out.nc,0,'Version','2.6')
  ncatt_put(out.nc,0,'Id',outFile)
  ncatt_put(out.nc,0,'Summary',paste('Global dataset of the Standardized
    Precipitation-Evapotranspiration Index (SPEI) at the ', sca,
    '-month', ifelse(sca==1,'','s'), ' time scale. ', comment, sep=''))
  ncatt_put(out.nc,0,'Keywords','drought, climatology, SPEI, Standardized
    Precipitation-Evapotranspiration Index')
  ncatt_put(out.nc,0,'Institution','Consejo Superior de Investigaciones
            Científicas, CSIC')
  ncatt_put(out.nc,0,'Url','http://sac.csic.es/spei')
  ncatt_put(out.nc,0,'Creators','Santiago Beguería <santiago.begueria@csic.es>
            and Sergio Vicente-Serrano <svicen@ipe.csic.es>')
  ncatt_put(out.nc,0,'Software','Created in R using the SPEI package
            (https://cran.r-project.org/web/packages/SPEI/;
            https://github.com/sbegueria/SPEI)')
  b <- match.call()
  ncatt_put(out.nc,0,'Call',
            paste(b[[1]],'(',names(b)[[2]],'=',b[[2]],', ',names(b)[[3]],'=',b[[3]],
                  ', ',names(b)[[4]],'=',b[[4]],')',sep=''))
  ncatt_put(out.nc,0,'Date',date())
  ncatt_put(out.nc,0,'Reference','Beguería S., Vicente-Serrano S., Reig F., Latorre B. (2014) Standardized precipitation evapotranspiration index (SPEI) revisited: Parameter fitting, evapotranspiration models, tools, datasets and drought monitoring. International Journal of Climatology 34, 3001-3023.')
  ncatt_put(out.nc,0,'Reference2','Vicente-Serrano S.M., Beguería S., López-Moreno J.I. (2010) A Multiscalar Drought Index Sensitive to Global Warming: The Standardized Precipitation Evapotranspiration Index. Journal of Climate 23, 1696–1718.')
  ncatt_put(out.nc,0,'Reference3','Beguería S., Vicente-Serrano S., Angulo-Martínez M. (2010) A multi-scalar global drought data set: the SPEIbase. Bulletin of the American Meteorological Society 91(10), 1351–1356.')
  ncatt_put(out.nc,'lon','long_name','longitude')
  ncatt_put(out.nc,'lon','limits',paste(out.nc$dim$lon$vals[1],
                                         out.nc$dim$lon$vals[out.nc$dim$lon$len],sep=', '))
  ncatt_put(out.nc,'lon','convention','Coordinates are referred to cell centres')
  ncatt_put(out.nc,'lat','long_name','latitude')
  ncatt_put(out.nc,'lat','limits',paste(out.nc$dim$lat$vals[1],
                                         out.nc$dim$lat$vals[out.nc$dim$lat$len],sep=', '))
  ncatt_put(out.nc,'lat','convention','Coordinates are referred to cell centres')
  ncatt_put(out.nc,'time','limits',paste(out.nc$dim$time$vals[1],
                                          out.nc$dim$time$vals[out.nc$dim$lat$len],sep=', '))

  # Dimensions
  dimlon <- out.nc.var$dim[[1]]$len
  dimlat <- out.nc.var$dim[[2]]$len
  dimtme <- out.nc.var$dim[[3]]$len

  # A vector with the number of days per month
  if (isSPEI) {
    ndays <- Hmisc::monthDays(as.Date(out.nc.var$dim[[3]]$vals,
                                      origin='1900-1-1'))
  }

  n <- block
  for (j in seq(dimlat,1,-n)-n+1) {
    start <- c(1,j,ifelse(is.na(tlapse[1]),1,tlapse[1]))
    count <- c(dimlon,n,dimtme)
    if (isSPEI) {
      x <- ncvar_get(etp.nc, varid=etp.nc$var[[1]]$name, start, count) * ndays
      x <- ncvar_get(pre.nc, varid=pre.nc$var[[1]]$name, start, count) - x
    } else {
      x <- ncvar_get(pre.nc, varid=pre.nc$var[[1]]$name, start, count)
    }
    if (isMask) {
      msk <- ncvar_get(mask.nc, 'mask', c(1,j), c(dimlon,n))
    }

    # Convert from matrix to list
    x.list <- vector('list', dimlon*n)
    for (i in 1:dimlon) {
      for (h in 1:n-1) {
        if (!isMask) {
          x.list[[h*dimlon+i]] <- x[i,{h+1},]
        } else {
          if (msk[i,{h+1}]==1) {
            x.list[[h*dimlon+i]] <- x[i,{h+1},]
          } else {
            x.list[[h*dimlon+i]] <- rep(NA,dimtme)
          }
        }
      }
    }

    # Compute series of anomalies
    speii <- function(d,s) {spei(d, s, na.rm=TRUE)$fitted}
    x.list <- sfLapply(x.list, speii, sca)

    # Convert back to matrix
    for (i in 1:dimlon) {
      for (h in 1:n-1) {
        part <- x.list[[h*dimlon+i]]
        part[is.nan(part/part)] <- NA
        x[i,{h+1},] <- part
      }
    }

    # Store into output netCDF file
    ncvar_put(out.nc, nam, x, c(1,j,1), count)
    nc_sync(out.nc)
  }

  # Close output netCDF file
  nc_close(out.nc)
}

sfInit(parallel=TRUE, cpus=4)
sfLibrary(SPEI)

for (i in c(1,3,6,12)) {
  cat(paste("Computing", i, "month SPEI...\n"))

  spei.nc(
    sca=i,
    inPre='./inputData/cru_ts4.03.1901.2018.pre.dat.nc',
    inEtp='./inputData/cru_ts4.03.1901.2018.pet.dat.nc',
    outFile=paste('./outputNcdf/spei',
                  formatC(i, width=2, format='d', flag='0'),'.nc',sep=''),
    title=paste('Global ',i,'-month',
                ifelse(i==1,'','s'),' SPEI, z-values, 0.5 degree',sep=''),
    comment='Using CRU TS 4.03 precipitation and potential evapotranspiration data',
    block=36,
    inMask=NA,
    tlapse=NA
  )

  cat(paste("Completed", i, "month SPEI!\n\n"))
}

sfStop()
print(list.files("./outputNcdf/"))
```

## South Africa Extraction

The `extract_sa_spei` function subsets each global netCDF to the bounding box of South Africa and returns a long-format data frame. Results are saved to `./south_africa_data/` for use in script 3.

```{r extract-sa}
# South Africa approximate boundaries
lon_min <- 16.5  # Western boundary
lon_max <- 32.9  # Eastern boundary
lat_min <- -34.8 # Southern boundary
lat_max <- -22.1 # Northern boundary

extract_sa_spei <- function(spei_file, time_scale) {

  nc <- nc_open(spei_file)

  lon  <- ncvar_get(nc, "lon")
  lat  <- ncvar_get(nc, "lat")
  time <- ncvar_get(nc, "time")

  dates   <- as.Date(time, origin = "1900-01-01")
  lon_idx <- which(lon >= lon_min & lon <= lon_max)
  lat_idx <- which(lat >= lat_min & lat <= lat_max)

  spei_data <- ncvar_get(nc, "spei",
                         start = c(min(lon_idx), min(lat_idx), 1),
                         count = c(length(lon_idx), length(lat_idx), -1))
  nc_close(nc)

  # Build one row per grid cell per time point
  df_list <- list()
  for (i in seq_along(lon_idx)) {
    for (j in seq_along(lat_idx)) {
      df_list[[length(df_list) + 1]] <- data.frame(
        longitude  = lon[lon_idx[i]],
        latitude   = lat[lat_idx[j]],
        date       = dates,
        spei       = spei_data[i, j, ],
        time_scale = time_scale
      )
    }
  }

  do.call(rbind, df_list)
}

spei_01 <- extract_sa_spei("./outputNcdf/spei01.nc", 1)
spei_03 <- extract_sa_spei("./outputNcdf/spei03.nc", 3)
spei_06 <- extract_sa_spei("./outputNcdf/spei06.nc", 6)
spei_12 <- extract_sa_spei("./outputNcdf/spei12.nc", 12)

if (!dir.exists("./south_africa_data")) dir.create("./south_africa_data")

write.csv(spei_01, "./south_africa_data/south_africa_spei_01month.csv", row.names = FALSE)
write.csv(spei_03, "./south_africa_data/south_africa_spei_03month.csv", row.names = FALSE)
write.csv(spei_06, "./south_africa_data/south_africa_spei_06month.csv", row.names = FALSE)
write.csv(spei_12, "./south_africa_data/south_africa_spei_12month.csv", row.names = FALSE)

# Summary statistics
for (ts in c(1, 3, 6, 12)) {
  df <- get(paste0("spei_", formatC(ts, width = 2, flag = "0")))
  cat(paste0("\n", ts, "-month SPEI:\n"))
  cat(paste("  Observations:", nrow(df), "\n"))
  cat(paste("  Grid cells:", length(unique(paste(df$longitude, df$latitude))), "\n"))
  cat(paste("  Date range:", min(df$date), "to", max(df$date), "\n"))
  cat(paste("  SPEI range:", round(min(df$spei, na.rm = TRUE), 2), "to",
            round(max(df$spei, na.rm = TRUE), 2), "\n"))
  cat(paste("  Mean SPEI:", round(mean(df$spei, na.rm = TRUE), 2), "\n"))
  cat(paste("  Missing values:", sum(is.na(df$spei)), "\n"))
}
```
